Evaluation of Cache-Oblivious Data Structures

### 2008-03-17 proposal/proposal.tex
Updated research proposal.

% TODO: abstract
%\begin{abstract}
%The cache-oblivious memory model is a model that allows algorithms and data structures to be evaluated in terms of memory transfers in a two-level memory hierarchy. Efficient data structurs and algorithms that operate on them are of interest because they are used in more complex, domain-specific algorithms.  This proposal suggests that emperical evidence is necessary to determine wether cache-oblivious data structures that have been described previously, perform well enough in practice so that they may replace traditional data structures.
%\end{abstract}

@@

It should be clear by now that a modern computer system lacks a single, central memory storage. Instead, there is a hierarchy of memory storage types, including at least the processor registers, L1/L2 cache, main memory and disk storage. At each of these levels, the next storage type is both significantly faster and significantly larger than the previous type. Typically, the processor can directly manipulate data contained in its registers only. To access data in a lower level in the memory hierarchy, the data must be transfered upward through the memory hierarchy. Memory is transfered in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).

@@

This raises the question whether cache-oblivious data structures are actually preferable to traditional data structures in practice. The theory suggests that cache-oblivious algorithms should perform well, but this can only be confirmed by evaluating their performance in a realistic environment and operating on a realistic data set. However, such empirical data is scarce when it comes to cache-oblivious data structures, as existing research has focused mainly on theoretical analysis. The goal of the proposed research project is to perform an empirical evaluation of previously described cache-oblivious data structures and compare them with their traditional counterparts (both cache-aware and cache-unaware data structures).

From the application programmer's point of view, data structures and algorithms can be grouped together based on the functionality they provide. For example, all comparison-based sorting algorithms can be considered to provide equivalent functionality, even if their performance is different. Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the project to data structures that provide similar functions. As will described below, we will limit ourselves to one particular kind of data structures, in the hope that our results may give some insight in the performance of cache-oblivious data structures in general.

@@

More specifically, these questions will be answered:
\newcounter{cnt}
\begin{list}{\arabic{cnt}.}{\usecounter{cnt}}
\item How do cache-oblivious data structures perform when compared to cache-aware data structures?
\item How do cache-oblivious data structures perform when compared to cache-unaware data structures?
\item How does the memory use of cache-oblivious data structures relate to the memory used by other data structures?
\end{list}
Because we are primarily interested in practical performance, the measure of performance will be the actual run-time of the experiments. Memory use is of interest as well (if only as a secondary criterion) as it may reveal that there is a trade-off between memory use and execution time of data structures.

\section{Research Approach}

% To add:
% - language of implementation
% - platform used
% - data used
% - how to test

The goal of the project is to evaluate a variety of data structures in a practical setting. For this, we need a realistic scenario that makes heavy use of data structures of which cache-oblivious implementations are available. A scenario that suits these needs, is that of a state space search algorithm (or, in short: state search). The purpose of such an algorithm is to find an optimal path from a (set of) given initial state(s) to a goal state. State search algorithms are often used to solve reachability problems, which occur for example in formal model checking. The main data structures used are a queue (either a regular queue or, if search heuristics are used, a priority queue) and a dynamic set. An example of a simple state search algorithm is given in program \ref{prog-search}.

@@

Since implementing a normal queue is fairly straightforward, the execution time of the algorithm will be dominated by the performance of the set, assuming domain-specific functions (like \verb#successors# and \verb#isGoal# in the sample code) are efficient. The main focus of the research project will therefore be on alternative implementations of the set data structure.

@@

\subsection{Set Operations}
A minimal set data structure should be able to determine whether or not a given element is a member of the set. A dynamic set differs from a static set in that it supports operations to add and/or remove elements from the set dynamically. In our experiments, we will start out with an empty set and only allow for insertion of new elements. This means the our dynamic set must support two basic operations:
\begin{itemize}
\item \verb#insert(S,X)# -- Inserts $X$ into set $S$, if it is not yet present.
\item \verb#contains(S,X)# -- Returns whether set $S$ contains $X$.
\end{itemize}
In practice, it is useful to be able to test if a set contains an element and add this element if it does not. This can be implemented by first testing for the existence, and then conditionally inserting the element (as is done in program \ref{prog-search}) but many data structures allow to combine this into a single operation. To take advantage of this possibility, we will require that the \verb#insert# operation also returns whether or not the element was newly inserted.

@@

\subsection{Measurement}
% TODO: which hardware and software platform used, et cetera. Do we discuss this?
As described in the problem statement, the main metric of interest is the total execution time of our algorithm. To be able to quantify the obtained results, we also need to be able to estimate the overhead incurred by the use of our testing framework and other outside influences. Two methods will be employed:
\begin{itemize}
\item Profiling tools such as OProfile \cite{oprofile} can be used to determine which part of the execution time was spent executing the testing framework, with a small overhead during execution.
\item A mock set implementation will be written that does not implement set functionality itself, but reads the answers to queries from a file. This can be used to give an upper bound on the overhead caused by the testing framework.
\end{itemize}

>>>>

% TODO: abstract?
%\begin{abstract}
%\end{abstract}

@@

It should be clear by now that a modern computer system lacks a single, central memory storage. Instead, there is a hierarchy of memory storage types, including at least the processor registers, L1/L2 cache, main memory and disk storage. At each of these levels, the next storage type is both significantly slower and significantly larger than the previous type. Typically, the processor can directly manipulate data contained in its registers only. To access data in a lower level in the memory hierarchy, the data must be transfered upward through the memory hierarchy. Memory is transfered in blocks of data of a fixed size (although bulk transfers involving multiple blocks of data at once are also supported at the memory and disk level).

@@

This raises the question whether cache-oblivious data structures are actually preferable to traditional data structures in practice. These data structures are designed to perform optimally in a model that was designed to more accurately reflect reality, which suggests they should perform well, but this can only be confirmed by evaluating their performance in a realistic environment and operating on a realistic data set. However, such empirical data is scarce, as existing research has focused mainly on theoretical analysis. The goal of the proposed research project is to perform an empirical evaluation of previously described cache-oblivious data structures and compare them with their traditional counterparts (both cache-aware and cache-unaware data structures).

From the application programmer's point of view, data structures and algorithms can be grouped together based on the functionality they provide. For example, all comparison-based sorting algorithms can be considered to provide equivalent functionality, even if their performance is different. Since it is not sensible to compare data structures which provide different functionality, it is necessary to limit the scope of the project to data structures that provide similar functions. As will be described below, we will limit ourselves to one particular kind of data structures, in the hope that our results may give some insight in the performance of cache-oblivious data structures in general.

@@

Because we are primarily interested in practical performance, actual execution time is the most important metric. Memory use is of interest as well (if only as a secondary criterion) as it may reveal that there is a trade-off between memory use and execution time of data structures. That means that the main question will be answered by first addressing the following questions:
\newcounter{cnt}
\begin{list}{\arabic{cnt}.}{\usecounter{cnt}}
\item How fast are cache-oblivious data structures when compared to cache-aware data structures?
\item How fast are cache-oblivious data structures when compared to cache-unaware data structures?
\item How much memory do cache-oblivious data structures use compared to other data structures?
\end{list}
These questions can be answered by implementing the relevant data structures and measuring their practical performance on a realistic data set.

\section{Research Approach}
The goal of the project is to evaluate a variety of data structures in a practical setting. For this, we need a realistic scenario that makes heavy use of data structures of which cache-oblivious implementations are available. A scenario that suits these needs, is that of a state space search algorithm (or, in short: state search). The purpose of such an algorithm is to find an optimal path from a (set of) given initial state(s) to a goal state. State search algorithms are often used to solve reachability problems, which occur for example in formal model checking. In a simple implementation, such as given in program \ref{prog-search}, the main data structures used are a queue and a dynamic set.

@@

Since implementing a normal queue is fairly straightforward, the execution time of the algorithm will be dominated by the performance of the set, assuming domain-specific functions (like \verb#successors# and \verb#isGoal# in the sample code) are efficient enough not to dominate the performance of the algorithm. The main focus of the research project will therefore be on alternative implementations of the set data structure.

@@

\subsection{Set Operations}
At a minimum, a set data structure should be able to determine whether or not a given element is a member of the set. A dynamic set differs from a static set in that it supports operations to add and/or remove elements from the set dynamically. In our experiments, we will start out with an empty set and only allow for insertion of new elements. This means the our dynamic set must support two basic operations:
\begin{itemize}
\item \verb#insert(S,X)# -- Inserts $X$ into set $S$, if it is not yet present.
\item \verb#contains(S,X)# -- Returns whether set $S$ contains $X$.
\end{itemize}
In practice, it is useful to be able to test if a set contains an element and add this element if it does not. This can be implemented by first testing for the existence, and then conditionally inserting the element (as is done in program \ref{prog-search}) but many data structures allow to combine this into a single operation. To take advantage of this possibility, we will require that the \verb#insert# operation also returns whether or not the element was newly inserted.

@@

\subsection{Measurement}
As described in the problem statement, the main metric of interest is the total execution time of our algorithm. To be able to quantify the obtained results, we also need to be able to estimate the overhead incurred by the use of our testing framework and other outside influences. Two methods will be employed:
\begin{itemize}
\item Profiling tools such as OProfile \cite{oprofile} can be used to determine which part of the execution time was spent executing the testing framework, with a small overhead during execution.
\item A mock set implementation will be written that does not implement set functionality itself, but reads the answers to queries from a file. This can be used to give an upper bound on the overhead caused by the testing framework.
\end{itemize}
